---
title: "Entry level R maps from African data | Cartes simples avec des données africaines" 
author: "afrimapr"
date: "07 July 2021"
output:
  xaringan::moon_reader:
    includes:
      after_body: insert-logo.html
    self_contained: true
    css: ["useR", "useR-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: center, middle

# Welcome!     |      Bienvenue!


---

## Meet your teaching team


_**English**_

**Instructors:** Andy South, Laurie Baker

**Helpers:** Anelda van der Walt, Shelmith Kariuki, Anne Treasure

<br>
<br>

_**French**_

**Instructor:** Ghislain Nono Gueye

**Helpers:** Ahmadou Dicko, Bryan Tegomoh

---

## Meet your fellow learners

```{r library packages, echo=FALSE, message=FALSE}

library(tmap) 
library(tidyverse)

```

```{r set tmap, echo=FALSE, message=FALSE}
tmap_mode("view") 
```

```{r learners map, fig.dim=c(4.8, 4.8), out.width="100%", echo=FALSE, message=FALSE, warning=FALSE}

# read in data
df <- read_csv("data/countries.csv")

## Mapping
# Count the number of registrations per country
my_countries <- df %>% 
   rename(name = "country") %>% 
   mutate(name = as_factor(name)) %>% 
   group_by(name) %>% 
   tally(name = "count")

# Load the World data set from tmap
data(World)

# Join the dataframe created above onto the World dataset to get country shapes
map_data <- World %>% left_join(my_countries, by = "name")

tm_shape(map_data) +
   tm_polygons("count", 
               palette="-magma", 
               contrast=.7, 
               id="name", 
               title="Number of participants")

```

---

## Collaborative document

Please note we have a _**shared google document**_ where you can ask questions, share ideas, etc. 

Please also take a minute to introduce yourselves here.

The link has been pasted in the Zoom chat channel:
https://docs.google.com/document/d/1nsD1pi-F88oJ8Dr2zA522G2liI5C1yQxL73QZVeZNWE

---

## Schedule

Topic |   | Duration
 :-- | --- | :-:
   |   | (minutes)
Welcome & Introduction  |   | 20
Navigating RStudio  |   | 15
Tutorial Session 1  |   | 50
Break  |   | 15
Tutorial Session 2  |   | 50
Break  |   | 15
Tutorial Session 3  |   | 50
Wrap up  |   | 5
Goodbyes  |   | 5

---

class: center, middle

# Session one

## 'Introduction to spatial data'

---

## What you will learn

- Loading R packages and data

- What is spatial data

- Creating static and interactive maps

- Overlaying several data types on a map

- Working with colour palettes on maps

---

## Example data we'll look at

- points (capital cities) 
- lines (highway network) 
- polygons (country boundaries) 
- gridded or raster data (population density)

```{r library_packages, echo=FALSE, message=FALSE}

#### LOAD PACKAGES AND DATA ----

library(sf) 
library(raster) 
library(afrilearndata)
library(tmap)
library(rgdal)
library(mapview)

```

```{r set tmap plot, echo=FALSE, message=FALSE}
tmap_mode("plot") 
```

```{r tmap-first-view, fig.height=3, dev='svg', echo=FALSE}

tmap::tm_shape(afripop2020) +
  tm_raster(palette = rev(viridisLite::magma(5)), breaks=c(0,2,20,200,2000,25000)) +
  tm_shape(africountries) +
  tm_borders("white", lwd = .5) +
  tm_shape(afrihighway) +
  tm_lines(col = "red") + 
  tm_shape(africapitals) +
  tm_symbols(col = "blue", alpha=0.4, scale = .6 )+
  tm_legend(show = FALSE)
```

---

### Example static maps we'll create

.pull-left[

```{r tmap-polygons-1c, fig.dim=c(4.8, 4.8), out.width="100%", echo=FALSE, message=FALSE}

tm_shape(africountries) +
  tm_polygons(col="income_grp", border.col = "white")

```

]

.pull-right[

```{r tmap-raster1b, fig.dim=c(4.8, 4.8), out.width="100%", echo=FALSE, message=FALSE}

tmap_mode('plot') 
tm_shape(afripop2020) +
  tm_raster(breaks=c(0,2,20,200,2000,25000))

```

]

---
### Example interactive maps we'll create


```{r set tmap plot2, echo=FALSE, message=FALSE}
tmap_mode("view") 
```

```{r tmap-interactive, fig.dim=c(4.8, 4.8), out.width="100%", echo=FALSE, message=FALSE, warning=FALSE}

tmap::tm_shape(afripop2020) +
  tm_raster(palette = rev(viridisLite::magma(5)), breaks=c(0,2,20,200,2000,25000)) +
  tm_shape(africountries) +
  tm_borders("white", lwd = .5) +
  tm_text("iso_a3", size = "AREA") +
  tm_shape(afrihighway) +
  tm_lines(col = "blue") + 
  tm_shape(africapitals) +
  tm_symbols(col = "green",  scale = .6 ) +
  tm_legend(show = FALSE)

```

---

class: center, middle

# Session two

## 'Reading my spatial data into R'

---

## What you will learn

- Read a variety of spatial data file types into R 

- Use `mapview` for quick interactive maps

- Options for making `mapview` maps more useful


---

## Interactive maps with `mapview`


```{r mapview-options-airports, fig.height=4, dev='svg', echo=FALSE, message=FALSE}

# 1. read into dataframe
filename <- system.file("extdata", "afriairports.csv", package="afrilearndata", mustWork=TRUE)
mydf <- readr::read_csv(filename)

mydf <- mydf[(1:100), ] #select first 100 rows just to make quicker online

# or can select a single country:
# mydf <- mydf[which(mydf$country_name == "Burkina Faso"), ]

# 2. convert to sf object
mysf <- sf::st_as_sf(mydf, 
                     coords=c("longitude_deg", "latitude_deg"),
                     crs=4326)

# 3. quick interactive plot
#mapview(mysf)
mapview(mysf, zcol='type', label='name', cex=2)   

```

---

class: center, middle

# Session three

## Practical

---

## What you will do

- Make a map from data of your own choice

- Learn where to find help 

- Share a screenshot or the code you used to create your map

---

## How this tutorial will work

Following this introduction, you will go to breakout rooms (English / French)
for the tutorial:

Topic |   | Duration
 :-- | --- | :-:
   |   | (minutes)
Navigating RStudio  |   | 15
Tutorial Session 1  |   | 50
Break  |   | 15
Tutorial Session 2  |   | 50
Break  |   | 15
Tutorial Session 3  |   | 50
Wrap up  |   | 5


Then back into the main training room for goodbyes

---

## Zoom

- After this introduction, when you are asked to join a break out room, click yes!

- At the end of the tutorial, the break out rooms will close and you will automatically be transferred back into the main room

- If you need help or have a question during the tutorial, please _**write a message in the zoom chat channel**_ - there are helpers and tutors in each break out room who will help you

- Please also note we have a _**shared google document**_ where you can ask questions, share ideas, etc. Please also take a minute to introduce yourselves here (the link has been pasted in the Zoom chat channel)
https://docs.google.com/document/d/1nsD1pi-F88oJ8Dr2zA522G2liI5C1yQxL73QZVeZNWE



---

## Code of Conduct

Please note that we will be following the useR!2021 Code of Conduct to provide a safe and inclusive conference experience for everyone regardless of age, gender, sexual orientation, disability, physical appearance, race, or religion (or lack thereof) (Details here: https://user2021.r-project.org/participation/coc/)

Anyone who violates this Code of Conduct will be reported to useR!2021, and removed from the tutorial.

---

class: center, middle

# afrimapr community meetups

---

## afrimapr community meetups

**Please join us!**

afrimapr has launched a monthly virtual meetup for people working on mapping African data in R. 

The community meets online on the 4th Wednesday of each month to discuss their work, solutions, ideas, resources, and more. 

To learn more or receive the calendar invitation with connection details, please join our afrimapr Google Group at https://groups.google.com/g/afrimapr/ and follow us on Twitter @afrimapr

---

## afrimapr

- Our website (http://afrimapr.org)

- Find us on GitHub (https://github.com/afrimapr)

- Follow us on Twitter (@afrimapr)

- Google group for discussion (https://groups.google.com/g/afrimapr)


